Two Problems for Sophistication

### 2015-01-20 Facticity/Facticity/paper.tex
intro tweaked

\title{Three Problems for Sophistication}

@@

\begin{enumerate}
\item Since it measures information, $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
For most proposed definitions of sophistication, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only strings that require an enormous amount of processing to construct, yield a large sophistication.

A valid definition of $S$ must contend with the following issues. First, the representation of the model is crucial. This is particularly important to those treatments that `open up' the definition of $K$, to consider the representation that witnesses the Kolmogorov complexity. In Section~\ref{section:indices}, we discuss the \emph{nickname problem}: the sophistication becomes highly dependent on the chosen programming language, unless it can represent functions with maximum efficiency.

The second issue is that of \emph{overfitting}, which we consider in Section~\ref{section:overfitting}. Overfitting is a common problem in the statistical literature, that refers to the tendency of model selection to choose a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. For example, in a polynomial regression setting, a polynomial of sufficiently high order can always cover all data points. However, if the data was sampled from a straight line, with a small amount of noise, such a polynomial would be far too complex: it would encode all noise inside the model. The same can occur for the sophistication. We somehow need to make sure that the two-part representation that determines the sophistication does not store noise in the ``structure'' part of the code. In statistics, this is typically addressed by penalising complex models. In sophistication, however, such penalties tend to break the balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{section:underfitting}) occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is a problem for sophistication because the considered models can be so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} can be described in a constant amount of bits, and  represent the data with an input of the length of the Kolmogorov complexity, essentially encoding all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions. While excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.

>>>>

\title{Two Problems for Sophistication}

@@

\begin{enumerate}
\item Since it measures information, $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold for sophistication: at the very least, there should be strict limits on the degree to which a change in the programming language can affect the sophistication.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language. Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $K(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, that are all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, that is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

For any string $x$, there may be many different two-part representations. The total code length can never be less than the Kolmogorov complexity, but it can come close. The balance between the two parts may vary; Figure~\ref{fig:todo} provides an illustration. The various proposals then define sophistication in terms of the length of the structure part of this two-part code for one of the representations that (almost) achieves the Kolmogorov complexity. However, For most of these definitions, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only strings that require an enormous amount of processing to construct, yield a large sophistication.

A valid definition of $S$ must contend with two important issues. First, the details of the way the model is encoded are important. There are two technically distinct approaches; in one of these one has to deal with the so-called ``nickname problem'' that strangely remains unresolved in several publications. These definitions yield a sophistication that is highly dependent on the chosen programming language, unless special care is taken, as discussed in Section~\ref{section:indices}.

The second issue is that of striking the right balance between underfitting and overfitting, which we consider in Section~\ref{section:balance}. Overfitting is a common problem in the statistical literature, that refers to the tendency of model selection to choose a very complex model that provides a very good fit to the observed data, but does not generalise well to unseen data. In the case of sophistication, overfitting occurs if the model that determines the sophistication contains much or even all of the noise. This is a real danger for sophistication, because the models are encoded so efficiently. In statistics, overfitting is often addressed by penalising complex models. In sophistication, however, such penalties tend to break the balance between structural information and noise, and lead to the opposite problem: underfitting.

Underfitting (Section~\ref{section:underfitting}) occurs when the selected model is simple, but fails to represent some of the structure present in the data. This is also a problem for sophistication because the considered models are so powerful. In particular, in any programming language, there are programs that implement an interpreter for \emph{another} language. Such \emph{universal models} can be described in a constant number of bits, and are able to reconstruct the data with an input of the length of the Kolmogorov complexity. Such a two-part representation essentially encodes all information as noise. If complex models are penalized, then the problem becomes to make sure that universal models are not \emph{always} preferred for complex data. The usual workaround is to restrict the set of allowed models to exclude the universal ones. For example, the model class may be restricted to the set of all total recursive functions. While this excludes universal models, it is questionable whether it adequately solves the problem of underfitting in general.
